\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{url}
\usepackage{amsmath,amssymb,amsfonts}

\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{float}
\usepackage{booktabs}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Operation Theater Planning Using Stochastic Programming and SAA\\
}

\author{\IEEEauthorblockN{1\textsuperscript{st} Jurgen van der Blom}
\IEEEauthorblockA{
j.j.vanderblom@student.utwente.nl}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Eric (H.Q.) van Ree}
\IEEEauthorblockA{
h.q.vanree@student.utwente.nl}
}

\maketitle

\input{sections/introduction}

\input{sections/problem_description}

\input{sections/literature_review}
%mathematical model
\input{sections/mathematical_model}

\input{sections/experimental_setup}

\input{sections/results_and_analysis}

\input{sections/blueprint_planning}

\input{sections/discussion}

\input{sections/conclusion}

\section{Author Statement}
The authors discussed the content of the report, reviewed the final version, and agreed to its submission.

Jurgen (50\%) Worked on the introduction, problem description, literature review including the SAA approach, and deterministic equivalent MILP. In addition, worked on the scenario generation distributions from the dataset by analysing the data. Together worked on the experimental setup and results of steps 1-4 (including VSS). Worked on the model extension for step 5 including the subgroup definition, and implemented it in code. Together worked on the discussion and conclusion.

Eric (50\%) Worked on the introduction, problem description, deterministic equivalent MILP, and recourse function. Furthermore implemented MILP in code, scenario sampling methods, symmetry breaking constraints, and warm-start solutions in code. Together worked on the experimental setup and results of steps 1-4 (including EVPI). Together worked on the discussion and conclusion.

\section{AI Usage Statement}
The group used Claude Sonnet 4.6 and Opus 4.8 during this assignment. It was used to assist with the code implementation, structuring the report, supporting the writing process, and evaluating the work. All AI-assisted output was critically assessed, and revised where necessary. The authors confirm that they fully understand the algorithms handed in and are able to explain all elements independently, in accordance with the course requirements regarding the permitted use of generative AI.

\input{sections/appendix}

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
