\section{Experimental setup and instance designs}
\label{sec:saa}

This section describes the experimental design used to find a suitable sampling
strategy and scenario count that balance outcome coverage with computational
tractability. We first describe how scenarios are constructed from the data,
then the three sampling strategies evaluated, and finally the overall
experimental design including parameter justification.

\subsection{Scenario Generation from Data}
\label{sec:scenario_data}

The dataset contains over $20{,}000$ patient treatment records, each including
specialty $q \in \{\text{CHI},\text{ORT},\text{KNO}\}$, surgery type, OR
identifier, waiting list identifier, and both planned and realised surgery
duration. In total, 186 unique surgery types across 12 ORs are recorded. For a
subset of records, the session identifier and within-session sequence are also
provided.

Two observations from the data motivate our scenario generation approach:
\begin{enumerate}
  \item Realised durations differ significantly across surgery types, making
        surgery type a strong predictor of duration.
  \item Planned durations outperform surgery-type historical averages as point
        estimates: the mean absolute error using planned durations is $12.8\%$
        lower than when using surgery-type averages.
\end{enumerate}

We therefore model the \emph{planned-duration error}
$\varepsilon_p = d_p^{\text{real}} - d_p^{\text{plan}}$ rather than the
realised duration directly, preserving the information content of the planned
duration. For each surgery type $\tau$, we estimate parameters
$\mu_\tau$ and $\sigma_\tau^2$ as the mean and variance of the observed errors
$\varepsilon_p$ over all historical patients of type $\tau$. Realised durations
are then sampled as:
%
\begin{equation}
  d_{ps} \sim
    \mathrm{LogNormal}\!\left(d_p^{\text{plan}} + \mu_\tau,\; \sigma_\tau^2
    \right),
  \qquad \forall\, p \in P,\; s \in S.
  \label{eq:duration_dist}
\end{equation}
%
Sampling the error rather than the raw duration ensures that surgery-type
averages do not absorb the additional predictive signal carried by the planned
duration. To generate a scenario $s$, one realisation $d_{ps}$ is drawn per
patient $p$ from \eqref{eq:duration_dist}. All patient durations within a
scenario are drawn independently; scenarios are also drawn independently of
each other.

To avoid sampling clinically implausible durations from the unbounded tails of
the lognormal distribution, each draw is truncated to the interval
$[\mu_p - 2\sigma_p,\, \mu_p + 2\sigma_p]$, with the lower bound additionally
floored at a small positive value to preclude non-positive durations. This
reflects the practical observation that a procedure is extremely unlikely to
take a drastically shorter or longer time than its planned duration suggests,
and that retaining such extreme outliers would distort both the SAA training set
and the out-of-sample evaluation. The same truncation is applied consistently to
all sampling methods and to the evaluation set, so the evaluation distribution
is the truncated lognormal that we treat as the operative ``true'' distribution.

\subsection{Sampling Strategies}
\label{sec:sampling}

We evaluate three sampling strategies to assess which provides the best
solution quality for a given sample size $N$.

\subsubsection{Random Sampling (RS)}
Each duration $d_{ps}$ is drawn independently and uniformly at random from
\eqref{eq:duration_dist}. Scenario probabilities are set to $\pi_s = 1/N$ for
all $s \in S$. RS is unbiased and asymptotically consistent, but with small
$N$ some regions of the distribution may be over- or under-represented, leading
to higher variance across replications.

\subsubsection{Latin Hypercube Sampling (LHS)}
LHS stratifies the marginal cumulative distribution of each patient's duration
into $N$ equally probable intervals $[(k-1)/N,\; k/N]$ for $k = 1, \dots, N$.
One quantile value is drawn uniformly from each interval per patient, and the
$N$ values are then randomly permuted across scenarios to form the scenario
matrix $d_{ps}$. Scenario probabilities are $\pi_s = 1/N$. By construction,
each marginal distribution is evenly covered, yielding lower variance than RS
for the same $N$~\cite{McKay1979}.

\subsubsection{Importance Sampling (IS)}

Importance Sampling (IS) is a variance reduction technique that samples more frequently from regions of the input space that would disproportionately influence the objective, and corrects for this bias through likelihood ratio reweighting. In
appointment scheduling, more extreme duration scenarios, where procedures run
substantially longer or shorter than expected, are the primary drivers of
overtime, idling, and patient waiting costs. Standard random sampling assigns
these tail scenarios a low probability, requiring a large number of scenarios to achieve reliable cost estimates.

To enrich the scenario set with such tail events, we apply a mean-shifting IS
strategy as used in \cite{aditri-2026}. Specifically, we partition the
scenario set $S$ into three equally sized subsets and draw each subset from a
shifted lognormal distribution with mean parameter $\mu_\tau - k\sigma_\tau$,
$\mu_\tau$, and $\mu_\tau + k\sigma_\tau$, respectively, where $k > 0$
controls the magnitude of the shift. The negative shift targets scenarios in
which procedures finish earlier than expected, increasing idling costs, while
the positive shift targets scenarios in which procedures overrun, increasing
overtime and downstream waiting costs.

To ensure that the SAA objective remains an unbiased estimator of the true
expected cost, each scenario $s$ is reweighted by its likelihood ratio under
the nominal distribution relative to the shifted distribution from which it
was drawn:
%
\begin{equation}
  \pi_s
    = \frac{\prod_{i} f(d_{is} \mid \mu_\tau,\, \sigma_\tau)}
           {\prod_{i} f(d_{is} \mid \mu_\tau + \delta_s \sigma_\tau,\, \sigma_\tau)},
  \label{eq:is_weights}
\end{equation}
%
where $f(\cdot \mid \mu, \sigma_\tau)$ denotes the nominal lognormal density,
$\delta_s \in \{-k,\, 0,\, k\}$ is the shift applied to the subset from
which scenario $s$ was drawn, and the product runs over all surgeries $i$ in
the schedule. Since surgery durations are sampled independently, the joint
probability of a scenario factorises into the product of the individual
marginal densities. Scenarios whose realised durations are consistent with the
nominal distribution receive a weight close to one, while scenarios drawn from
a strongly shifted distribution receive lower weight in proportion to how far
they deviate from the nominal mean.

A direct consequence of this reweighting is that the enriched tail scenarios —
precisely those that drive overtime and cascade waiting — carry comparatively
\emph{small} importance weights, since they are unlikely under the nominal
distribution. The effective sample size, $N_{\text{eff}} = 1 / \sum_s \pi_s^2$,
is therefore smaller than $N$, and for small $N$ the SAA objective under IS may
require \emph{more} scenarios than RS to reach a stable estimate of the expected
cost. The compensating benefit is robustness: because IS guarantees that
worst-case (long-duration) realisations are represented in every sample, the
resulting policies tend to hedge more conservatively against tail events. We
return to this trade-off when interpreting the convergence results in
Section~\ref{sec:results}.

\subsection{Experimental Setup}
\label{sec:exp_setup}

\paragraph{Evaluation sample.}
Solution quality is assessed on an out-of-sample evaluation set of size
$N' = 1{,}000$ scenarios generated by RS from the treatment duration distribution
\eqref{eq:duration_dist}. Because the evaluation set is large, RS provides
near-complete coverage of the outcome space and serves as an approximation for the true
expected cost.

\paragraph{SAA parameters.}
We use $M = 10$ independent replications per sample size $N$. The sample size is
swept from an initial $N = 10$ in steps of $5$. Rather than estimating the true
optimum $v^*$ to bound the optimality gap, we monitor the dispersion of the
out-of-sample objective across replications through a $95\%$ confidence interval
on the out-of-sample mean cost:
%
\begin{equation}
  \bar{v} \pm z_{0.975}\,\frac{\hat{\sigma}_M}{\sqrt{M}},
  \label{eq:ci}
\end{equation}
%
where $\bar{v}$ and $\hat{\sigma}_M$ are the sample mean and standard
deviation of the $M$ out-of-sample evaluations, and $z_{0.975} = 1.96$.
For each sample size we record this interval and plot it against $N$; the
appropriate $N$ is then selected \emph{post hoc} from the convergence plot as
the point beyond which the mean cost stabilises and the interval no longer
narrows materially.

\paragraph{Expected performance of sampling strategies.}
We expect LHS to outperform RS for small $N$ due to its stratification
guarantee, and IS to excel when the objective is dominated by high-duration
scenarios (overtime, late-cascade waiting times). RS, while unbiased, is
expected to require the largest $N$ for comparable variance. These hypotheses
are tested by applying each method across Steps~1--4 and comparing out-of-sample
confidence intervals.

\paragraph{Experiments}
For Steps~1 and~2, we use the patient set from waiting list~4, which contains patients
planned across 6 unique OR sessions, each with a known session identifier and
within-session position label. This instance is used for both the timing-only model
(Step~1) and the timing-and-sequencing model (Step~2), with the session and position
labels serving to fix the assignment and sequencing in Step~1.

For Step~3, four patient sets, drawn from waiting lists~4, 7, 10, and~13, are
used to assess solution quality across different patient populations.
Each waiting list contains patients planned across at least 6 unique sessions.

Each of the four samples is solved using the SAA framework with all three sampling strategies.
Starting from $N = 10$ scenarios and increasing in steps of 5, we apply.

For each sample set, $M = 10$ independent replications are solved and the resulting
out-of-sample objective values are evaluated on a fixed evaluation set of
$N' = 1{,}000$ RS scenarios which we generate once per sample. Convergence is monitored via the $95\%$ confidence
interval~\eqref{eq:ci}, which is kept for each $(sample,sampling~approach,N)$ and used to plot convergence and evaluate when $N$ is large enough for each of the sampling approaches (for each of the samples).
