\section{Results and Analysis}
\label{sec:results}
\subsection{Expected Value of Perfect Information}

The Expected Value of Perfect Information (EVPI) quantifies the expected gain from knowing
the exact scenario before making any decisions. It is estimated as follows.

First, the stochastic MILP is solved using $S$ training scenarios drawn from the
chosen sampling distribution, yielding a here-and-now policy
$(X^*, Y^*, A^*, D^*)$.
Second, $I$ independent evaluation scenarios are generated.
For each evaluation scenario $i$, two objective values are computed:
\begin{itemize}
    \item $z_{\text{policy}}(i)$: the cost using the policy trained on S evaluated on scenario $i$.
    \item $z_{\text{perf}}(i)$: the cost with perfect information, obtained by solving the MILP with scenario $i$ as the sole scenario, so all first-stage decisions can be optimised with full knowledge of the realised durations.
\end{itemize}
The value of perfect information for scenario $i$ is
\[
    \text{VPI}(i) = z_{\text{policy}}(i) - z_{\text{perf}}(i) \;\geq\; 0,
\]
and the EVPI is estimated as the sample mean over the $I$ evaluation scenarios:
\[
    \widehat{\text{EVPI}} = \frac{1}{I} \sum_{i=1}^{I} \text{VPI}(i).
\]

Table~\ref{tab:vpi} presents the per-scenario costs for 10 training scenarios
and 10 evaluation scenarios. In 8 out of 10 evaluation scenarios,
$z_{\text{perf}}(i) = 0$, meaning that a penalty-free schedule exists when
the surgery durations are known in advance. For these scenarios, the entire
cost incurred by the policy is attributable to uncertainty, and
$\text{VPI}(i) = z_{\text{policy}}(i)$. Only 2 scenarios (i = 1 and i = 4)
require penalties even under perfect information, suggesting that these
particular realisations of surgery durations are inherently infeasible to
schedule without cost regardless of the decision rule applied. The estimated
expected value of perfect information over the 10 evaluation scenarios is
$\widehat{\text{EVPI}} = 158.05$, indicating that on average a large share of
the policy cost could in principle be eliminated by resolving duration
uncertainty prior to scheduling.

\begin{table}[h]
    \centering
    \caption{Per-scenario costs and Value of Perfect Information (VPI).}
    \label{tab:vpi}
    \begin{tabular}{rrrr}
        \toprule
        $i$ & $z_{\text{policy}}(i)$ & $z_{\text{perf}}(i)$ & $\text{VPI}(i)$ \\
        \midrule
         1 & 225.146 & 16.048 & 209.098 \\
         2 & 119.461 &  0.000 & 119.461 \\
         3 & 119.849 &  0.000 & 119.849 \\
         4 & 177.055 &  2.742 & 174.313 \\
         5 & 153.461 &  0.000 & 153.461 \\
         6 & 128.878 &  0.000 & 128.878 \\
         7 & 140.100 &  0.000 & 140.100 \\
         8 & 186.231 &  0.000 & 186.231 \\
         9 & 174.386 &  0.000 & 174.386 \\
        10 & 174.679 &  0.000 & 174.679 \\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Value of the Stochastic Solution}

The Value of the Stochastic Solution (VSS) measures the benefit of explicitly
accounting for duration uncertainty rather than planning against a single
deterministic forecast. It is defined as
\[
    \text{VSS} = \text{EEV} - \text{RP},
\]
where RP (the \emph{recourse problem}) is the expected out-of-sample cost of the
stochastic (SAA) policy, and EEV (the \emph{expected result of using the
expected value solution}) is the expected out-of-sample cost of the
deterministic policy.

The estimation procedure consists of three steps.
First, the \emph{expected value} (EV) policy is obtained by solving the MILP with
a single scenario in which every surgery takes its expected duration; this yields
a here-and-now schedule $(X^{\text{EV}}, Y^{\text{EV}}, A^{\text{EV}}, D^{\text{EV}})$.
Second, the stochastic (SAA) policy is obtained by solving the MILP with $N$
training scenarios drawn from the chosen sampling distribution, yielding
$(X^{\text{SAA}}, Y^{\text{SAA}}, A^{\text{SAA}}, D^{\text{SAA}})$.
Third, both fixed first-stage policies are evaluated on the \emph{same}
independent out-of-sample set of $N'$ scenarios, generated with a seed distinct
from those used to train either policy. Sharing a single evaluation set
eliminates evaluation-sample noise from the difference, so that
$\text{VSS} \geq 0$ holds up to solver optimality gaps rather than being
polluted by two independent sample means. The two out-of-sample means are
\[
    \text{EEV} = \frac{1}{N'} \sum_{i=1}^{N'} z\!\left(\text{EV policy}, i\right),
    \qquad
    \text{RP} = \frac{1}{N'} \sum_{i=1}^{N'} z\!\left(\text{SAA policy}, i\right),
\]
where $z(\cdot, i)$ is the realised cost of a fixed first-stage policy under
evaluation scenario $i$, computed by the recourse simulation. A large VSS
indicates that incorporating uncertainty into the planning model yields a
substantial cost reduction over the naive deterministic plan.

% Results to be added.